
# OLS Refresher and Problems

This notebook refreshes by illustration what ordinary least squares (OLS) is, its principal problems, and why Econometricians discovered quasiexperimental methods.
Throughout, I will maintain the assumption that readers are familiar with the potential outcomes model of causality (i.e., the [Rubin-Neyman Model](https://en.wikipedia.org/wiki/Rubin_causal_model)) because classical quasiexperimental econometrics is foundationally based on the POM.

```{r}
library(tidyverse)
library(stargazer)
library(MASS)
source("sim_functions.r")
set.seed(202307)
n <- 1000
reps <- 1000
```

## Monte Carlo Experiments

Before we talk about the properties of OLS, we need to talk about _how_ such properties are actually investigated empirically.

To investigate the properties of statistical estimators, you _cannot_ actually run tests on real-world data.
The problem with that approach is that in most cases, you _do not know_ the ground truth.
This is especially true in the Social Sciences, where the true values of many model parameters are continuing topics of heated debate
(e.g., look up the ongoing debate on elasticities of demand in various industries in Economics).

Instead, in theoretical Statistics, the gold-standard for investigating the empirical properties of estimators is the Monte Carlo study.
A Monte Carlo study is a data simulation exercise.
A typical study procedure follows the following algorithm (in pseudocode):
```
results_list = repeat(0, n_iterations)
for (i in n_iterations) {
    data <- data_generating_process(true_model_parameters)
    estimator <- estimator_class(estimator_hyperparameters)
    results <- estimator.fit(data)
    results_list[i] <- results.parameter_estimates
}
display(summary(results_list))
```
The goals of this study procedure are simple:

1. Is the estimator unbiased ($\mathbb{E}[\beta^{estimator}]=\beta^{true}$)?
2. Is the estimator consistent ($\beta^{estimator}\underset{n\rightarrow\infty}{\rightarrow}\beta^{true}$)?
3. How large is the empirical standard error?
4. Under what DGP conditions are the estimator unbiased? Consistent? Efficient?

As described, a Monte Carlo study requires 3 elements:

1. A DGP.
2. An estimator.
3. Computational power.

In this case, (2) and (3) are assumed given so we discuss (1).
When studying linear estimators, in general we need not discsus non-linear DGPs.
Linear estimators generally fail in these cases.

### The Linear DGP

The Linear DGP simulates a generic simultaneous equations model:
\[
    \begin{align*}
        y_{i,1} &= \mu_{y,1} + \gamma_{y,1,1} y_{i,1} + \cdots + \gamma_{y,m,1} y_{i,m} + \gamma_{x,1,1} x_{i,1} + \cdots + \gamma_{x,k,1} x_{i,k} + \xi_{i,1} \\
        y_{i,2} &= \mu_{y,2} + \gamma_{y,1,2} y_{i,1} + \cdots + \gamma_{y,m,2} y_{i,m} + \gamma_{x,1,2} x_{i,1} + \cdots + \gamma_{x,k,2} x_{i,k} + \xi_{i,2} \\
        & \vdots \\
        y_{i,m} &= \mu_{y,m} + \gamma_{y,1,m} y_{i,1} + \cdots + \gamma_{y,m,m} y_{i,m} + \gamma_{x,1,m} x_{i,1} + \cdots + \gamma_{x,k,m} x_{i,k} + \xi_{i,m} \\
        x_{i,1} &= \mu_{x,1} + \epsilon_{i,1} \\
        x_{i,2} &= \mu_{x,2} + \epsilon_{i,2} \\
        & \vdots \\
        x_{i,k} &= \mu_{x,k} + \epsilon_{i,k}
    \end{align*}
\]

Which in matrix notation is more simply
\[
    \begin{align*}
        \mathbf{Y}_{i} &= \mu_{Y} + \Gamma_{Y} \mathbf{Y}_{i} + \Gamma_{X} \mathbf{X}_{i} + \xi_{i} \\
        \mathbf{X}_{i} &= \mu_{X} + \epsilon_{i} \\
        Cov\left(\begin{bmatrix}\epsilon_{i}\\\xi_{i}\end{bmatrix}\right) &= \Sigma \\
        m &\ge 1 \\
        k &\ge 0
    \end{align*}
\]
Where $m,k$ are the lengths of column vectors $\mathbf{Y}_{i},\mathbf{X}_{i}$ respectively.
This generic DGP encompasses most common cases:

1. The single-equation case is parameterised by $m=1,\gamma_{y,1,1}=0$.
2. The $1$-factor model case is a case of a multivariate simple linear model parameterised by $m>1,\Gamma_{Y}=\mathbf{0},k=1$.
3. The $k$-factor model with non-overlapping factors case is parameterised by $m>1,\Gamma_{Y}=\mathbf{0},k>1$, and $\Gamma_{X}$ block diagonal.
4. The simultaneity case is parameterised by $m>1,|\Gamma_{Y}|\ne0$.

Additionally, this DGP is simple to generate:
\[
    \begin{align*}
        \mathbf{Y}_{i} &= \left( \mathbf{I} - \Gamma_{Y} \right)^{-1} \left( \mu_{Y} + \Gamma_{X} \mathbf{X}_{i} + \xi_{i} \right) \\
        \mathbf{X}_{i} &= \mu_{X} + \epsilon_{i}
    \end{align*}
\]

We can now describe the code for this DGP:
```
function(n, mu_y, mu_x, sigma_xe, gamma_y, gamma_x) {

    #   Count m, k
    n_y <- length(mu_y)
    n_x <- length(mu_x)

    #   Ensure mu_y, mu_x are column vectors
    mu_y <- matrix(mu_y, ncol = 1)
    mu_x <- matrix(mu_x, ncol = 1)

    #   Jointly generate xi, x (x = mu_x + epsilon)
    mu_x <-  rbind(mu_x, matrix(rep(0, n_y), ncol = 1))
    x_xi <- mvrnorm(n, mu_x, sigma_xe)

    #   Assign xi, x
    x <- matrix(x_xi[, 1:n_x], ncol = n_x)
    xi <- matrix(x_xi[, (n_x + 1):(n_x + n_y)], ncol = n_y)

    #   Compute mu_y + gamma_x * x + xi
    y <- x %*% gamma_x + xi
    y <- sweep(t(y), 1, mu_y, FUN = "+")

    #   Solve Y in (I - gamma_y) * Y = mu_y + gamma_x * x + xi
    y <- solve(diag(n_y) - gamma_y, y)

    #   Export as tibble
    rownames(y) <- paste("y", 1:nrow(y), sep = "")
    colnames(x) <- paste("x", 1:ncol(x), sep = "")
    y <- cbind(t(y), x)
    y <- as_tibble(y)
    return(y)
}
```

In practice, for maintainability and portability, this function has been saved in a separate file and loaded as a library.
We can now briefly demonstrate this DGP in action:
```{r}
data <- sim_eq_data(
    n = n,
    mu_y = c(0),
    mu_x = c(0),
    sigma_xe = diag(2),
    gamma_y = matrix(c(0), ncol = 1),
    gamma_x = matrix(c(1), ncol = 1)
)
plot(data$x1, data$y1)
```

The actual implementation of this DGP is slightly more involved than described above to allow for non-Normal joint distributions through a simple implementation using the [Probability Integral Transform](https://en.wikipedia.org/wiki/Probability_integral_transform) with [Inverse Transform Sampline](https://en.wikipedia.org/wiki/Inverse_transform_sampling).
Note that this while this method is not theoretically sound, the alternative of using [Copulas](https://en.wikipedia.org/wiki/Copula_(probability_theory)) is less straightforward to understand and detracts from the main purpose of allowing for Non-Normal joint distributions.
Below, the distribution of the errors of $y$ have been changed to a $U(-1,1)$ distribution.

```{r}
data <- sim_eq_data(
    n = n,
    mu_y = c(0),
    mu_x = c(0),
    sigma_xe = diag(2),
    gamma_y = matrix(c(0), ncol = 1),
    gamma_x = matrix(c(1), ncol = 1),
    eps_tnsfmrs = c(function(v) qunif(pnorm(v), -1, 1))
)
plot(data$x1, data$y1)
```

A second difference in this implementation is to permit generating indicator variables by providing threshold values.

```{r}
data <- sim_eq_data(
    n = n,
    mu_y = c(0),
    mu_x = c(0),
    sigma_xe = diag(2),
    gamma_y = matrix(c(0), ncol = 1),
    gamma_x = matrix(c(1), ncol = 1),
    alpha_x = c(0)
)
plot(data$x1, data$y1)
```

While this simple DGP has generally good properties, its limitations includes a lack of the following features:

1. Interaction and higher-order polynomial terms.
2. Time series generation.

However, since it is not currently of urgent importance to have these features, how to generate these variables in a generalisable vectorised manner will be left for a different time and as an exercise to the reader in the meantime.

## OLS Capabilities

Using the Monte Carlo method with the linear DGP, we can now investigate the properties of the Ordinary Least Squares (OLS) estimator.
To refresh, recall the Gauss-Markov Theorem.
Under the following assumptions, OLS is the Best Linear Unbiased Estimator (BLUE):

1. $y_{i} = \alpha + \beta x_{i} + \epsilon_{i}$.
2. $\mathbb{E}\epsilon_{i} = 0$.
3. $Cov(x_{i}, \epsilon_{i}) = 0$ or $\mathbb{E}[\epsilon \mid \mathbf{X}] = 0$.
4. $\epsilon_{i} \sim \mathcal{N}(0,\sigma_{i})$.
5. $\sigma_{i} = \sigma \forall i$.

Assumption (1) is necessary for linearity.
Assumptions (2) and (3) are necessary for consistency and unbiasedness (specifically, uncorrelatedness only yields consistency while mean independence yields unbiasedness).
Assumptions (4) and (5) are necessary for efficiency (best estimator).

As a reminder, the OLS point estimator is robust to violations of assumptions (4) and (5) but not (1) or (3).
Violations of assumptions (4) and (5) will require a heteroskedasticity-robust (sandwich) estimator for the standard errors while violations of assumptions (1) or (3) render OLS unusable.
Assumption (2) is trivially satisfied when the DGP and model contain a mean ($\alpha$).

### Univariate Case

First, to see that OLS recovers the true parameters of a system when the model approximates the DGP sufficiently closely, observe the classical univariate case:
\[
    y_{i} = \alpha + \beta x_{i} + e_{i}\\
    corr(x_{i}, e_{i}) = 0
\]
OLS is unbiased and consistent if its basic assumptions are satisfied.

```{r}
data <- sim_eq_data(
    n = n,
    mu_y = c(0),
    mu_x = c(0),
    sigma_xe = diag(2),
    gamma_y = matrix(c(0), ncol = 1),
    gamma_x = matrix(c(1), ncol = 1)
)
lm_res <- lm(y1 ~ x1, data = data)
stargazer(lm_res, type = "text")
```

### Monte Carlo

With a MC design, we can visualise the spread of the coefficient estimates.

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_y = c(0),
        mu_x = c(0),
        sigma_xe = diag(2),
        gamma_y = matrix(c(0), ncol = 1),
        gamma_x = matrix(c(1), ncol = 1)
    )
    lm_res <- lm(y1 ~ x1, data = data)
    coefs[i] <- lm_res$coefficients["x1"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    True Value = 1
    "
)
```

### Multiple Regressors Case

OLS is able to isolate the marginal contribution of a single covariate with multiple regressors even when regressors are correlated.
\[
    y_{i} = \alpha + \mathbf{x}_{i}^{\prime} \beta + e_{i}\\
    corr(\mathbf{x}_{i}, e_{i}) = \mathbf{0}
\]
OLS is still able to recover the true marginal effect of a covariate with correlated covariates, as long as the correlation is not too high to create a collinear system.
Uncertainty is proportional to correlation across covariates because it becomes increasingly difficult for OLS to partition the variance of the dependent variable between covariates as covariates increasingly resemble a single underlying factor with independent measurement error.

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_y = c(0),
        mu_x = c(0, 0),
        sigma_xe = rbind(
            c(1, .7, 0),
            c(.7, 1, 0),
            c(0, 0, 1)
        ),
        gamma_y = matrix(c(0), ncol = 1),
        gamma_x = matrix(c(1, -1), ncol = 1)
    )
    lm_res <- lm(y1 ~ x1 + x2, data = data)
    coefs[i] <- lm_res$coefficients["x2"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    Marginal Effect of One Covariate (True Value = -1)
    "
)
```

## Robustness to Some Violations of Assumptions

Some assumptions concern only the efficiency of OLS under the Gauss-Markov Theorem (i.e., OLS is no longer BLUE without these assumptions).
However, since consistency and biasedness are not affected, these assumptions do not violate the ability of OLS to recover the true parameters.

### Violating Gaussian-Distribution of Errors

While most applied statisticians are taught the Standard OLS, it is a well-known fact that the Normal distribution of errors is not required for consistency or unbiasedness.
In fact, Normality is only a necessary condition under the Maximum Likelihood derivation of the OLS estimator, which yields OLS as the best (most efficient) unbiased estimator.
Under the Method of Moments approach, the necessary assumptions are
\[
    \mathbb{E}[\epsilon_{i}] = 0 \\
    \mathbb{E}[X_{i}\epsilon_{i}] = \mathbf{0}
\]
In which case, observe that
\[
    \mathbb{E}[X_{i}(Y_{i}-X_{i}^{\prime}\beta)] = \mathbf{0} \\
    \Rightarrow \beta = \left(\mathbb{E}[X_{i}X_{i}^{\prime}]\right)^{-1} \mathbb{E}[X_{i}Y_{i}]
\]
Which is the OLS solution.

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_y = c(0),
        mu_x = c(0),
        sigma_xe = diag(2),
        gamma_y = matrix(c(0), ncol = 1),
        gamma_x = matrix(c(1), ncol = 1),
        eps_tnsfmrs = c(function(v) qunif(pnorm(v), -1, 1), list(NULL))
    )
    lm_res <- lm(y1 ~ x1, data = data)
    coefs[i] <- lm_res$coefficients["x1"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    True Value = 1
    "
)
```

What differs in this case is the solution for the standard error of the estimator.

### Non-Continuous Covariates

Covariates also do not need to be continuous under OLS.
This should be clear from the Method of Moments estimator for OLS.
Below we demonstrate this point.

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_y = c(0),
        mu_x = c(0),
        sigma_xe = diag(2),
        gamma_y = matrix(c(0), ncol = 1),
        gamma_x = matrix(c(1), ncol = 1),
        alpha_x = c(1)
    )
    lm_res <- lm(y1 ~ x1, data = data)
    coefs[i] <- lm_res$coefficients["x1"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    True Value = 1
    "
)
```


## Non-Exogenous Regressors

### Univariate Case

Among the OLS assumptions, one strictly cannot be violated: independence.
With a simple univariate case, we show what happens when the covariate is endogenous (e.g., correlated with the unobserved error):
\[
    y_{i} = \alpha + \beta x_{i} + e_{i}\\
    corr(x_{i}, e_{i}) \ne 0
\]

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_x = c(0),
        mu_y = c(0),
        sigma_xe = rbind(
            c(1, .5),
            c(.5, 1)
        ),
        gamma_y = matrix(c(0), ncol = 1),
        gamma_x = matrix(c(-1), ncol = 1)
    )
    lm_res <- lm(y1 ~ x1, data = data)
    coefs[i] <- lm_res$coefficients["x1"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    True Value = -1
    "
)
```

### Endogeneity Bias

The true parameter value is `-1` but we will reject $H_{0}: \beta=-1$ with p-value approaching zero.
The severity of the bias depends on two parameters:

1. The coefficient of the object in the error term that the endogenous variable is correlated with.
2. The coefficient of correlation between the object in the error term and the endogenous variable.

\[
    \beta^{bias} = \beta^{true} + \gamma \frac{cov(x_{i},w_{i})}{var(x_{i})}
\]
Where $w_{i}$ is the component of the error that is correlated with the endogenous variable $x_{i}$ and $\gamma$ is the true marginal effect of $w_{i}$ on the outcome $y_{i}$.

### Omitted Variables Bias

To illustrate this more clearly, we showcase the effect of the bias with the classic omitted relevant variables case with structural model
\[
    y_{i} = \alpha + \beta x_{i} + \gamma z_{i} + \epsilon_{i} \\
    corr(x_{i}, \epsilon_{i}) = 0 \\
    corr(z_{i}, \epsilon_{i}) = 0 \\
    corr(x_{i}, z_{i}) \ne 0
\]
and estimated model
\[
    y_{i} = a + b x_{i} + e_{i}
\]
In the linear DGP system this is parameterised by generating the system
\[
    \begin{align*}
        y_{1,i} &= \alpha_{1} + \beta y_{2,i} + \gamma_{1} x_{i} + \epsilon_{1,i} \\
        y_{2,i} &= \alpha_{2} + \gamma_{2} x_{i} + \epsilon_{2,i} \\
        x_{i} &= \mu + \xi_{i}
    \end{align*}
\]

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    data <- sim_eq_data(
        n = n,
        mu_y = c(0, 0),
        mu_x = c(0),
        sigma_xe = diag(3),
        gamma_y = rbind(
            c(0, -1),
            c(0, 0)
        ),
        gamma_x = matrix(c(4, 1), nrow = 1)
    )
    lm_res <- lm(y1 ~ y2, data = data)
    coefs[i] <- lm_res$coefficients["y2"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "OLS Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of OLS Coefficient Estimates
    True Value = -1
    "
)
```

In this case, the true coefficient is `-1` but OLS estimates a positive coefficient (`-1+4*(1/2)=1`).

### Other Sources of Endogeneity

Endogeneity results from many other sources such as but not limited to:

1. Non-Random Selection into Treatment.
2. Reverse Causation.
3. Simultaneity.
4. Classical Measurement Error in Covariates.
5. Truncation and Censoring.

## Intuition for Quasiexperiment Methods

In the presence of such data quality problems, which are relatively common (e.g., non-complete controls, observational data of general equilibrium systems, measurement errors), how do we really go about obtaining causal estimates?
In Economics, the disciplinary consensus is to use quasiexperimental methods (sometimes also microeconometric methods, to distinguish from what macroeconomists normally do).
At the fundamental level, this is a set of 4 methods:

1. Instrumental Variables (IV).
2. Difference-in-Difference (DID).
3. Regression Discontinuity Design (RDD).
4. Propensity Score Matching (PSM).

All advanced econometric methods (e.g., synthetic control, Event Study, Tobit/Heckit, Arellano-Bond, Robust/Doubly-Robust Machine Learning, et cetera) are fundamentally level some variation of one of these 4 methods.
What are the intuition for these methods?
In short:
If you could not run your own controlled trial (typically, the system is not ethically amenable to controlled experimentation), look for a naturally-conducted trial (i.e., naturally-occuring exogenous variation).
Each of these 4 methods target an intuitive source of exogenous variation:

1. IV: A determinant of the endogenous variable that itself does not directly enter the system being studied itself; intuitively, in controlled trials, assignment to treatment is the instrument (assignment $\rightarrow$ covariate of interest $\rightarrow$ outcome)
2. DID: Selection into treatment is by outcome level but not outcome growth rate.
3. RDD: Selection into treatment is by outcome level but subjects around the assignment cut-off point are basically identical due to natural exogenous variation in outcome levels.
4. PSM: Selection Bias is entirely captured by the probability of being selected for treatment and not the characteristics leading to treatment.

The appropriate application of these methods allow economists to infer causality from purely observational data.
However, the price of these methods is highly limited external validity in many cases.
The other notebooks will go through IV, DID, and RDD.
