
# Basics of Difference-in-Difference

## Intuition

Many interesting social phenomena and questions have a pretest-posttest flavour.
For example, did the influx of refugees into Germany in 2015 reduce German natives' wage growth?
Did the expansion of Career Conversion Programmes in Singapore increase local firms' productivity?

Difference-in-Differences (DID) is both a method and conceptual framework for approaching causal inference in non-controlled pretest-posttest settings.
First, to understand why DID is necessary (especially in Economics, where it was first popularised), it is necessary to understand when other na&iuml;ve treatment effects estimators fail.
Using the Rubin-Neyman Potential Outcomes Model (POM), we define:

- $D_{i}=0,1$ as the treatment assignment of $i$ in period $1$.
- $t=0,1$ as the periods (before and after).
- $Y_{i,t}(d)$ as the potential outcome of $i$ in period $t=0,1$ given assignment $d=0,1$.
- $Y_{i,t}$ as the actual realised outcome of $i$ in period $t=0,1$.
- $\Delta_{i,t}=Y_{i,t}(1)-Y_{i,t}(0)$ as the Treatment Effect for $i$ in period $t=0,1$.

With pretest-posttest settings, there are two na&iuml;ve methods to estimate the Average Treatment Effect (ATE).

1. The Before-After Estimator:
    \[
        \begin{align*}
            \theta^{BA} &= \mathbb{E}[Y_{i,1} - Y_{i,0} \mid D_{i}=1] \\
            &= \mathbb{E}[Y_{i,1}(1) - Y_{i,0}(0) \mid D_{i}=1] \\
            &= \mathbb{E}[Y_{i,1}(1) - Y_{i,1}(0) + Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=1] \\
            &= \mathbb{E}[\Delta_{i,1} \mid D_{i}=1] + \mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=1]
        \end{align*} 
    \]
2. The Treatment-Control Estimator:
    \[
        \begin{align*}
            \theta^{TC} &= \mathbb{E}[Y_{i,1} \mid D_{i}=1] - \mathbb{E}[Y_{i,1} \mid D_{i}=0] \\
            &= \mathbb{E}[Y_{i,1}(1) \mid D_{i}=1] - \mathbb{E}[Y_{i,1}(0) \mid D_{i}=0] \\
            &= \mathbb{E}[Y_{i,1}(1) - Y_{i,1}(0) \mid D_{i}=1] \\
            &\quad + \mathbb{E}[Y_{i,1}(0) \mid D_{i}=1] - \mathbb{E}[Y_{i,1}(0) \mid D_{i}=0] \\
            &= \mathbb{E}[\Delta_{i,1} \mid D_{i}=1] \\
            &\quad + \mathbb{E}[Y_{i,1}(0) \mid D_{i}=1] - \mathbb{E}[Y_{i,1}(0) \mid D_{i}=0]
        \end{align*} 
    \]

In both cases, it is clear that in observational data, the additional term is generally non-zero.
In the first case, $\mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=1]$ is the time trend.
In the second case, $\mathbb{E}[Y_{i,1}(0) \mid D_{i}=1] - \mathbb{E}[Y_{i,1}(0) \mid D_{i}=0]$ is the selection bias.

Experimentalists will recognise that selection bias is eliminated in randomised controlled trials (RCTs) because under random assignment, $D_{i} \perp Y_{i,t}(0), Y_{i,t}(1)$.
However, note that causal inference methods are in general for situations where RCTs were _not_ possible.
In the absence of RCTs, DID can be applied if the Common Trends Assumption (CTA) is reasonable:
\[
    \mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=1] = \mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=0]
\]
Under this assumption, observe:
\[
    \begin{align*}
        \theta^{DID} &= \mathbb{E}[Y_{i,1} - Y_{i,0} \mid D_{i}=1] - \mathbb{E}[Y_{i,1} - Y_{i,0} \mid D_{i}=0] \\
        &= \mathbb{E}[\Delta_{i,1} \mid D_{i}=1] \\
        &\quad + \mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=1] \\
        &\quad - \mathbb{E}[Y_{i,1}(0) - Y_{i,0}(0) \mid D_{i}=0] \\
        &= \mathbb{E}[\Delta_{i,1} \mid D_{i}=1]
    \end{align*} 
\]
The DID yields exactly the ATE of the treated group in the posttest (after) period.

## Before-After Estimator bias

```{r}
library(tidyverse)
library(stargazer)
library(MASS)
set.seed(202307)
n <- 100
reps <- 100
```

DID DGP requirements are not difficult and as such complex functions will not be used.

```{r}
e <- mvrnorm(n = 2 * n, mu = c(0, 0), Sigma = diag(2))
x <- matrix(e[, 1], ncol = 1)
e <- matrix(e[, 2], ncol = 1)
t <- matrix(c(rep(0, n), rep(1, n)), ncol = 1)
d_assign <- matrix(rep(1 * (x[1:n,] > 0), 2), ncol = 1)
d <- 1 * (t == 1) * (d_assign == 1)
y_pom_0 <- 3 * (d_assign == 1) + 1 * (t == 1) + e
y_pom_1 <- y_pom_0 + 2
y <- y_pom_0 * (d == 0) + y_pom_1 * (d == 1)
data <- tibble(t, d_assign, d, y_pom_0, y_pom_1, y, x, e)
data
```

The DGP has been made unnecessarily verbose to make clear what the true parameters of this system are.
In this setting, the true ATE is $2$.

We can now investigate what the BA estimator returns:
```{r}
stargazer(lm(y ~ t, data = filter(data, d_assign == 1)), type = "text")
```

The estimator seems centred on $3$, which is not surprising since the ATE is $2$ and the time trend is $1$.
Note that we are not controlling for $x$ as in most observational data where DID is used, the running variable that codetermines the potential outcome and treatment assignment is unknown.

## Treatment-Control Estimator bias

For completeness' sake, we can also observe what the TC estimator returns:
```{r}
stargazer(lm(y ~ d_assign, data = filter(data, t == 1)), type = "text")
```

This is also not surprising, since the ATE is $2$ and the selection bias is $3$.

## The DID Estimator

Given the introduction, it should not be surprising that DID resolves this issue and generates an unbiased estimate of the ATE.
Note that although the DID is presented as a Method of Moments estimator, in practice it is easily estimated by a saturated linear model.

```{r}
stargazer(lm(y ~ d_assign + t + I(d_assign*t), data = data), type = "text")
```

Or in Monte Carlo format, 

```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    e <- mvrnorm(n = 2 * n, mu = c(0, 0), Sigma = diag(2))
    x <- matrix(e[, 1], ncol = 1)
    e <- matrix(e[, 2], ncol = 1)
    t <- matrix(c(rep(0, n), rep(1, n)), ncol = 1)
    d_assign <- matrix(rep(1 * (x[1:n,] > 0), 2), ncol = 1)
    d <- 1 * (t == 1) * (d_assign == 1)
    y_pom_0 <- 3 * (d_assign == 1) + 1 * (t == 1) + e
    y_pom_1 <- y_pom_0 + 2
    y <- y_pom_0 * (d == 0) + y_pom_1 * (d == 1)
    data <- tibble(t, d_assign, d, y_pom_0, y_pom_1, y, x, e)
    lm_res <- lm(y ~ d_assign + t + I(d_assign*t), data = data)
    coefs[i] <- lm_res$coefficients["I(d_assign * t)"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "DID ATE Point Estimates",
    ylab = "Density",
    main = "
    Distribution of DID ATE Estimates
    True Value = 2
    "
)
```

## Extensions

Like IV, DID is a family of methods.
Of all the extensions, two in particular stand out for their wide usage:

1. Event Study: This method conceptually extends DID to the case of multiple pre- and post-test periods.
    Suppose the existence of a longer panel with time periods $t=h,h+1,\cdots,-1,0,1,\cdots,k$ where $t$ denotes "periods after treatment" (i.e., $t=0$ is the period when treatment is applied).
    In this case, the usual procedure is to use $t=-1$ as the base case and estimate
    \[
        y_{i,t} = \alpha + \beta D_{i} + \sum_{\tau=h,\tau\ne-1}^{k} \gamma_{\tau} \mathbf{1}\{\tau=t\} +  \sum_{\tau=h,\tau\ne-1}^{k} \delta_{\tau} \mathbf{1}\{\tau=t\} \times D_{i} + \epsilon_{i,t}
    \]
    Which yields $\delta_{\tau}$ as the ATE in period $\tau$ for the treated group.

    For $\tau<-1$, estimates of $\delta_{\tau}$ serve as a proxy test of the CTA as well.
    If $\delta_{\tau}\ne0$ for any $\tau<-1$, this calls into question the CTA.
2. Triple Difference: This method conceptually extends DID into the realm of heterogeneous treatment effects.
    Suppose that treatment effects differ by gender.
    Then, similar to the Event Study case, we set one gender as the base case (typically "male") and estimate
    \[
        \begin{align*}
            y_{i,t} &= \alpha + \beta D_{i} + \gamma t + \sum_{g \in \mathcal{G}} \eta_{g} \mathbf{1}\{gender_{i}=g\} \\
            &\quad + \delta D_{i} t + \sum_{g \in \mathcal{G}} \pi_{g} D_{i} \mathbf{1}\{gender_{i}=g\} + \sum_{g \in \mathcal{G}} \sigma_{g} t \mathbf{1}\{gender_{i}=g\} \\
            &\quad + \sum_{g \in \mathcal{G}} \Delta_{g} t D_{i} \mathbf{1}\{gender_{i}=g\}+ \epsilon_{i,t}
        \end{align*}
    \]
    In which case $\delta + \Delta_{g}$ comprise the ATE for $g$.
    Note that gender is merely being used as an example here and any group by which ATEs are heterogeneous may be used.

## Repeated Cross-Section Data

A notable implication of DID being a Method of Moments estimator is that actual longitudinal data is not a requirement.
Specifically, repeated cross-sections may also be used, as long as the sample in all periods are representative of the same underlying population and the treated group is identifiable in all periods.
In fact, one of the first studies using DID actually used repeated cross-section data:

[Card, D. (1990). The impact of the Mariel boatlift on the Miami labor market.](https://doi.org/10.1177/001979399004300205)

## Summary

In observational data, treatments cannot be evaluated using a na&iuml;ve pretest-posttest methodology despite the presence of a control group due to unobserved time trends and selection bias.
Instead, Econometricians investigated the DID method to help practitioners perform causal inference in such situations.

In this notebook, I briefly demonstrate the canonical case where both the before-after and treatment-control estimators are biased and inconsistent and show that DID is unbiased and consistent under the same circumstances.
I further discuss two conceptual extensions to multiple time periods and heterogeneous ATEs and one use-case extension to repeated cross-section data, for which the actual experiment to prove the unbiasedness and consistency of DID under these settings are left as a future exercise.
