
# Basics of Random Discontinuity Design

## Intuition

Many research questions of interest in the Social Sciences revolve around the effect of being subject to certain treatments.
Often, these treatments either cannot be feasibly or ethically controlled, or they have occurred in the past without any controlled rollout.
In the Economics of Education, certain questions of interest that fall into this category include:

1. What is the value and benefit of holding academically weak students for a year?
2. What is the marginal value of attending a higher-ranked high school or college?
3. What is the marginal benefit of going to vocational school over college, or vice versa?

In many of these questions, the common underlying pattern is that there is clear selection into treatment where the expected treatment effect of the treatment is expected to be highly correlated with the selection criteria.
For example, on average we expect that attending a higher-ranked college to yield higher returns in the future (i.e., increase lifetime earnings).
However, selection into higher-ranked colleges is dependent in "ability", which is also expected to yield higher returns (i.e., *ceteris paribus*, a "more able" child will have greater lifetime earnings than a "less able" child).

So how do we disentangle such effects, given that we cannot ethically "randomly assign students to colleges"?
In the vast majority of cases, we usually cannot and trying to collect better data in the future is the only solution.
However, under certain circumstances we can exploit random variation in human behaviour and performance to study those at the margins of the treatment assignment threshold--this is the Random Discontinuity Design (RDD) method.

RDD uses the fact that at any arbitrary threshold based on certain behaviours and performance, people that fall onto either side of that threshold are in fact quite similar.
E.g., does there exist any real difference between a child who scored 49 and failed a subject (and therefore may need remedial or to retake that class) and a child who scored 51 and passed?
As is intuitive, the RDD methodology is about restricting our attention to only those observations that fall within a sufficiently small bandwidth around the treatment assignment threshold.

We demonstrate with a MC experiment.

```{r}
library(tidyverse)
library(stargazer)
library(MASS)
set.seed(20230703)
n <- 100000
reps <- 100
```

## Homogeneous Treatment Effect

The simplest DGP is a straightforward linear model with higher-order polynomial terms without requiring any endogeneity or correlatedness among covariates or heterogeneous ATE.

```{r}
e <- mvrnorm(n = n, mu = c(0, 0, 0, 0), Sigma = diag(4))
z <- matrix(e[, 1], ncol = 1)
x <- matrix(e[, 2], ncol = 1)
xi <- matrix(e[, 3], ncol = 1)
e <- matrix(e[, 4], ncol = 1)
r <- z + .25 * xi
d <- 1 * (r > 0)
y <- x - 5 * z ^ 3 + 2 * d + e
data <- tibble(y, d, x, r)
data
```

First, we show that the unbiased Average Treatment Effect (ATE) can still be recovered with a sufficiently high polynomial order in the running variable.

```{r}
stargazer(lm(y ~ x + d + r + I(r^2) + I(r^3), data = data), type = "text")
```

However, note that this strategy assumes that the marginal effect of the running variable follows a smooth curve that can be approximated with a sufficiently high polynomial order.

```{r}
stargazer(lm(y ~ x + d + r, data = data), type = "text")
```

Alternatively, under the RDD intuition, we can simply recover the true ATE in a local area around the treatment assignment threshold by analysing only observations in a small bandwidth around that threshold.

```{r}
stargazer(lm(y ~ x + d, data = filter(data, -.1 < r & r < .1)), type = "text")
```

In Monte Carlo format,
```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    e <- mvrnorm(n = n, mu = c(0, 0, 0, 0), Sigma = diag(4))
    z <- matrix(e[, 1], ncol = 1)
    x <- matrix(e[, 2], ncol = 1)
    xi <- matrix(e[, 3], ncol = 1)
    e <- matrix(e[, 4], ncol = 1)
    r <- z + .25 * xi
    d <- 1 * (r > 0)
    y <- x - 5 * z ^ 3 + 2 * d + e
    data <- tibble(y, d, x, r)
    lm_res <- lm(y ~ x + d + r, data = filter(data, -.1 < r & r < .1))
    coefs[i] <- lm_res$coefficients["d"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "RDD Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of RDD Coefficient Estimates
    True Value = 2
    "
)
```

If you play around with the simulation parameters, you will notice that the smaller the bandwidth used (and the greater the number of observations simulated to compensate), the less biased the estimator.

## Heterogeneous Treatment Effect

In most cases, the ATE estimated can only be interpreted as a Local ATE (LATE), meaning that there is no external validity to this result.
The ATE estimated is strictly only for those observations that fall right around the threshold.
For an intuition, refer to Figure 2 of [Imbens and Lemieux (2007)](https://doi.org/10.1016/j.jeconom.2007.05.001) which illustrates this point.

What does this mean in practice?
This implies that for example one cannot run a RDD on the effect of just marginally being assigned to additional remedial lessons and assume that _all_ students benefit by the same quantum if they were also assigned to those lessons.
We can demonstrate with a simple example.
```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    e <- mvrnorm(n = n, mu = c(0, 0, 0, 0), Sigma = diag(4))
    z <- matrix(e[, 1], ncol = 1)
    x <- matrix(e[, 2], ncol = 1)
    xi <- matrix(e[, 3], ncol = 1)
    e <- matrix(e[, 4], ncol = 1)
    r <- z + .25 * xi
    d <- 1 * (r > 0)
    y <- x - 3 * z ^ 3 + 2 * d - d * z + e
    data <- tibble(y, d, x, r)
    lm_res <- lm(y ~ x + d + r, data = filter(data, -.05 < r & r < .05))
    coefs[i] <- lm_res$coefficients["d"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "RDD Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of RDD Coefficient Estimates
    True Value = 2
    "
)
```

In this case, the treatment effect is still $2$ around the threshold, but now observe what happens if the threshold shifts:
```{r}
coefs <- rep(0, reps)
for (i in 1:reps) {
    e <- mvrnorm(n = n, mu = c(0, 0, 0, 0), Sigma = diag(4))
    z <- matrix(e[, 1], ncol = 1)
    x <- matrix(e[, 2], ncol = 1)
    xi <- matrix(e[, 3], ncol = 1)
    e <- matrix(e[, 4], ncol = 1)
    r <- z + .25 * xi
    d <- 1 * (r > 1)
    y <- x - 3 * z ^ 3 + 2 * d - d * z + e
    data <- tibble(y, d, x, r)
    lm_res <- lm(y ~ x + d + r, data = filter(data, 0.95 < r & r < 1.05))
    coefs[i] <- lm_res$coefficients["d"]
}
hist(
    coefs,
    freq = FALSE,
    breaks = 25,
    xlab = "RDD Coefficient Point Estimates",
    ylab = "Density",
    main = "
    Distribution of RDD Coefficient Estimates
    True Value = 1
    "
)
```

In general, the greatest drawback of the RDD method is the inability to convincingly generalise its findings despite being the cleanest causal identification method in the Econometrics toolkit.

## Fuzzy Thresholds

It is not always the case that the threshold is as clean as simulated here.
When the thresold is unclear and the probability of assignment to treatment resembles a sigmoid curve rather than a sharp discontinuous break, this is known as a Fuzzy RDD.

As would be intuited, Fuzzy RDD is handled by first estimating the probability of assignment to treatment and then using that probability of assignment in the place of the treatment assignment variable in the main equation.

## Summary

Regression Discontinuity Design is a causal inference method investigated by Econometricians to handle the case where the method of treatment assignment is known to follow a running variable.
Normal statistical test procedures do not normally work on such settings due to clear selection bias.
However, the presence of a running variable permits the Econometrician to pursue an alternative approach of estimating the ATE only locally around the assignment threshold.

The intuition is that in many cases, the running variable comprises some true underlying state with some measurement error.
Hence, observations in the vicinity of the threshold are otherwise identical and comparable save for being assigned to the treatment or not due to chance.

While threats to identification are not formally discussed, it suffices for the reader to understand intuitively that RDD only works if the running variable cannot be manipulated to fall just above or below the threshold.
For example, if we observe that the distribution of children's grades shows an excess of "just-passing" grades, then manipulation around the threshold has occurred and RDD is no longer a consistent estimator of the Local ATE.
